---
title: "The Secret Weapon: section 10.9 of Regression and Other Stories"
date: "2024-10-27"
categories: [ROS]
---


In section 10.9 of Regression and Other Stories, the authors introduce the idea of fitting the same model to many data sets. They begin the section by saying, "it is common to fit a regression model repeatedly...to subsets of an existing data set." But then two paragraphs later call this idea a "secret weapon" because "it is so easy and powerful but yet is rarely used." So which is it? Common or rarely used? 

Anyway, here's [the code](https://github.com/avehtari/ROS-Examples/blob/master/NES/nes_linear.R) they use to demonstrate this idea using NES data. This creates Figure 10.9 on page 149. While I enjoyed working through this code and figuring out how it works, I decided to re-implement it using data frames and ggplot2.

First I load the {rstanarm} package and read in the data.

```{r message=FALSE}
library(rstanarm)
data <- read.table("https://github.com/avehtari/ROS-Examples/raw/refs/heads/master/NES/data/nes.txt")
```

Next I modify their custom function as follows. The changes are in the `coefs` object that is created. Instead of a vector, I return a data frame. I also add the variable name and year to the data frame.

```{r}
coef_names <- c("Intercept", "Ideology", "Black", "Age_30_44", 
                "Age_45_64", "Age_65_up", "Education", "Female", "Income")
regress_year <- function (yr) {
  this_year <- data[data$year==yr,]
  fit <- stan_glm(partyid7 ~ real_ideo + race_adj + factor(age_discrete) +
                    educ1 + female + income,
                  data=this_year, warmup = 500, iter = 1500, refresh = 0,
                  save_warmup = FALSE, cores = 1, open_progress = FALSE)
    coefs <- data.frame(var = coef_names, 
                      coef = coef(fit), 
                      se = se(fit), 
                      year = yr)
}
```

Now I run the function using `lapply()`, combine the list of data frames into one data frame, add upper and lower limits of a 50% confidence interval, and set the variable name as a factor.

```{r}
sum2 <- lapply(seq(1972,2000,4), regress_year)
sumd <- do.call(rbind, sum2)
sumd$upper <- sumd$coef + sumd$se*0.67
sumd$lower <- sumd$coef - sumd$se*0.67
sumd$var <- factor(sumd$var, levels = coef_names)
```

And finally I create the plot using {ggplot2}.

```{r}
library(ggplot2)
ggplot(sumd) +
  aes(x = year, y = coef) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = lower, ymax = upper), width = 0) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ var, scales = "free") +
  theme_classic()
```

According to the book, "ideology and ethnicity are the most important, and their coefficients have been changing over time."

The above approach uses Bayesian modeling via the {rstanarm} package. We can also deploy the secret weapon using a frequentist approach via the `lmList()` function from the {lme4} package. Below I first subset the data to select only the columns I need for the years 1972 - 2000. This is necessary due to the amount of missingness in the data.

```{r message=FALSE}
library(lme4)
vars <- c("partyid7", "real_ideo" , "race_adj" , "age_discrete" ,
  "educ1" , "female" , "income", "year")
d <- data[data$year %in% seq(1972,2000,4),vars]
fm1 <- lmList(partyid7 ~ real_ideo + race_adj + factor(age_discrete) +
                educ1 + female + income | year, 
              data = d)
```

Calling summary on the object lists a summary of coefficients over time. 

```{r}
summary(fm1)
```

To create the plot, I need to do some data wrangling. Below I extract the coefficients from the summary, which returns an array. I then use the `adply()` function from the {plyr} package to convert the array to a data frame. Then I add year, upper 50% CI limit, and lower 50% CI limit to the data frame. I also change the variable column to a factor so the order of the coefficients will be preserved in the plot.

```{r}
sout <- summary(fm1)$coefficients

library(plyr)
sumd2 <- adply(sout, .margins = 3, .id = "Var")
sumd2$year <- seq(1972,2000,4)
sumd2$upper <- sumd2$Estimate + sumd2$`Std. Error`*0.67
sumd2$lower <- sumd2$Estimate - sumd2$`Std. Error`*0.67
sumd2$Var <- factor(sumd2$Var, labels = coef_names)
```

And once again I create the plot.

```{r}
ggplot(sumd2) +
  aes(x = year, y = Estimate) +
  geom_point() +
  geom_errorbar(mapping = aes(ymin = lower, ymax = upper), width = 0) +
  geom_hline(yintercept = 0, linetype = 2) +
  facet_wrap(~ Var, scales = "free") +
  theme_classic()
```

The result is almost identical to the one created using {rstanarm}.