---
title: "The null distribution of deviance"
date: "2026-03-23"
categories: [logistic regression]
---

In their book [Data Analysis Using Regression and Multilevel/Hierarchical Models](https://sites.stat.columbia.edu/gelman/arm/), Gelman and Hill state that adding a predictor to a GLM model that doesn't improve fit will only decrease deviance by about 1. That is also the mean of a chi-square distribution with one degree of freedom, which is the null distribution of deviance between two models with no difference in fit. Let's simulate this.

Below I simulate two sets of data from a standard normal distribution and generate a binomial outcome using the `rbinom()` function. Notice I use the `plogis()` function to make the logit transformation on the linear predictor. Also notice the linear predictor only uses x1. Then I fit two binary logistic regression models: one with just x1 and one with x1 and x2.

```{r}
set.seed(1)
n <- 500
x1 <- rnorm(n)
x2 <- rnorm(n)
lp <- 0.2 + 0.5*x1
y <- rbinom(n = n, size = 1, prob = plogis(lp))
m1 <- glm(y ~ x1, family = binomial)
m2 <- glm(y ~ x1 + x2, family = binomial)
```

Since x2 does not improve model fit, we expect the difference in residual deviance between the two models to be about 1.

```{r}
cbind(m1 = deviance(m1), m2 = deviance(m2))
```

The difference in residual deviance can be tested with the `anova()` function, which you can use when one model is nested within another. In this case m1 is nested within m2 since we can imagine m1 being the same as m2 with a 0 coefficient for the x2 predictor. 

```{r}
anova(m1, m2)
```

The deviance listed is 1.1079. The null distribution for deviance when two models are no different and only differ by one coefficient is a chi-square distribution with 1 degree of freedom. Based on the small deviance and large p-value, we fail to reject the null hypothesis of no difference in the fit of the models. 

Let's re-run the simulation above 1000 times and look at the distribution of deviances. The `replicate()` function makes quick work of this. Notice I save the result of the call to `anova()` to extract the deviance, which is in the second position of the "Deviance" element.

```{r}
rout <- replicate(n = 1000, expr = {
  n <- 500
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  lp <- 0.2 + 0.5*x1
  y <- rbinom(n = n, size = 1, prob = plogis(lp))
  m1 <- glm(y ~ x1, family = binomial)
  m2 <- glm(y ~ x1 + x2, family = binomial)
  out <- anova(m1, m2)
  out$Deviance[2]})
```

As expected, the mean of the deviances is about one.

```{r}
mean(rout)
```

Now plot a histogram of the deviances and overlay a chi-square distribution with one degree of freedom. (I used trial-and-error to determine the value of `n`, which is the number of values used to evaluate the chi-square function.) The deviances are indeed well-modeled with a chi-square distribution.

```{r}
hist(rout, freq = F, breaks = 50, 
     main = "Distribution of binomial deviances")
curve(expr = dchisq(x, df = 1), from = 0, to = 20, 
      add = TRUE, col = "red", n = 401, lwd = 2)
```

If you prefer a ggplot version, we can do the following. (Again, I had to use trial and error to get the `xlim` bounds just right.)

```{r}
library(ggplot2)
d <- data.frame(dev = rout)
ggplot(d) +
  aes(x = dev) +
  geom_density() +
  geom_function(fun = dchisq, args = list(df = 1), color = "red", 
                xlim = c(0.2, 20))
```


We can also demonstrate this using poisson models.

```{r}
rout2 <- replicate(n = 1000, expr = {
  n <- 500
  x1 <- rnorm(n)
  x2 <- rnorm(n)
  lp <- 0.2 + 0.5*x1
  y <- rpois(n = n, lambda = exp(lp))
  m1 <- glm(y ~ x1, family = poisson)
  m2 <- glm(y ~ x1 + x2, family = poisson)
  out <- anova(m1, m2)
  out$Deviance[2]})

mean(rout2)
hist(rout2, freq = F, breaks = 50, 
     main = "Distribution of poisson deviances")
curve(expr = dchisq(x, df = 1), from = 0, to = 20, add = TRUE, col = "red", n = 401, lwd = 2)
```

To learn how model residual deviance is calculated, see my [UVA Library StatLab article](https://library.virginia.edu/data/articles/understanding-deviance-residuals). 